Staged Abstract Interpreters

### 2019-03-25 paper/20_prelim.tex
section 2

\paragraph{Basic type definitions} We start with some basic type definitions.
The identifiers in the program are represented by strings. The two default
components in the interpreter are environments @Env@ and stores @Store@, i.e.,
mappings from identifiers to addresses and mappings from addresses to values,
respectively. But the @Addr@ and @Value@ are just declared as abstract types.

@@

\paragraph{Binding-time abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. If we instantiate @R@ as an identity
type (i.e., @type R[T] = T@), then the generic interpreter will execute the program.
In Section \ref{stagedinterp}, we will instantiate @R@ using LMS's built-in
next-stage type annotation @Rep@, which makes the interpreter act as a compiler.

\paragraph{Monadic operations} We define the return type of the interpreter as
@Ans@, which is an arbitrary monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain constraints have to be posed onto the type @AnsM@. Here, we use a
structural type @MonadOps@ to require @AnsM@ to at least implement @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time; accordingly the outer @R@ defined in the trait is passed
to @MonadOps@. Inside of @MonadOps@, @R[_]@ wraps the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When acting as
compilers, we will also replace the monads to the ones that work on staged
values.

@@

\paragraph{Primitive operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier, and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is also wrapped
by @R[_]@. The other one would simply wrap the address with our monadic type
@AnsM[_]@.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the interpreter.
The method @num@ and @close@ deal with primitive values, which lift literal
terms (e.g., lambdas) to our value representation (e.g., closures).
The method @get@ simply retrieves the value mapped from the identifier @x@ in
the environment and store. Conditionals and arithmetics is handled by @br0@
and @arith@, respectively. The methods @ap_clo@ takes a function value and an
argument value and then does the application.

@@

\paragraph{The interpreter} Now we can define the semantics-agnostic interpreter
in monadic form. The essential idea is to traverse the abstract syntax tree
while maintaining the effects such as reader and state. 
It is worth noting that the interpreter is written in open-recursive style -- it
can not refer to itself directly, instead, @eval@ takes an additional parameter
@ev@ of type @Expr => Ans@ to refer to itself. Accordingly, the method @close@
that lifts lambda terms to closures and @ap_clo@ that applies functions also
takes an extra @ev@.

@@

Now we can instantiate a concrete interpreter. We first concretize the abstract
types as follows. Since the interpreter is concrete, for the purpose that
properly models the heap, the type \texttt{Int} is used for address space
\texttt{Addr}. The environment and store are just ordinary \texttt{Map}s in
Scala. A value can be either a tagged number \texttt{NumV} or a closure
\texttt{CloV} that contains a lambda term and an environment.

\begin{lstlisting}
trait Concrete extends Semantics {
  type Addr = Int;   sealed trait Value
  case class CloV($\lambda$: Lam, $\rho$: Env) extends Value
  case class NumV(i: Int) extends Value
  type Env = Map[Ident, Addr];  type Store = Map[Addr, Value]
}
\end{lstlisting}

We proceed to implement the components that define the semantics. As mentioned
before, type \texttt{R} is concretized as an identity type function, thus all
types are just ordinary types.

\begin{lstlisting}
object ConcInterp extends Concrete {
  type R[+T] = T
  val $\rho$0 = Map[Ident,Addr]();  val $\sigma$0 = Map[Addr,Value]()
  def get($\rho$: Env, x: Ident): Addr = $\rho$(x)
  def put($\rho$: Env, x: Ident, a: Addr): Env = $\rho$ + (x -> a)
  def get($\sigma$: Store, a: Addr): Value = $\sigma$(a)
  def put($\sigma$: Store, a: Addr, v: Value): Store = $\sigma$ + (a -> v)
  def alloc($\sigma$: Store, x: Ident): Addr = $\sigma$.size + 1
  def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Env): Value = CloV($\lambda$, $\rho$)
  def num(i: Lit): Value = NumV(i.i)
  def apply_closure(ev: EvalFun)
    (f: Value, arg: Value, $\sigma$: Store): Ans = f match {
      case CloV(Lam(x, e), $\rho$) =>
        val $\alpha$ = alloc($\sigma$, x); val $\rho$_* = put($\rho$, x, $\alpha$)
        val $\sigma$_* = put($\sigma$, $\alpha$, arg); ev(e, $\rho$_*, $\sigma$_*)
    }
  def branch0(cnd: Value, thn: => Ans, els: => Ans): Ans = 
    cnd match { case NumV(i) => if (i == 0) thn else els }
  def prim_eval(op: Symbol, v1: Value, v2: Value): Value = 
    op match {
      case '+ => v1.asInstanceOf[NumV].i + v2.asInstanceOf[NumV].i
      ...
    }
}
\end{lstlisting}

Finally, \texttt{get} and \texttt{put} directly manipulate the map of
environment and store. An allocation function is used when we need to update the
store: in the concrete case, \texttt{alloc} always returns a fresh address of
the store. \texttt{close} and \texttt{num} lift syntactic literals to values.
\texttt{apply\_closure} takes the function value, argument value and the latest
store, then extends the environment and store after allocating a fresh address,
and continue evaluating the body \texttt{e} of the lambda term.

>>>>

\paragraph{Basic Types} We start with some basic type definitions used in the
interpreter. The identifiers in the program are represented by strings. The two
default components in the interpreter are environments @Env@ and stores @Store@,
i.e., mappings from identifiers to addresses and mappings from addresses to
values, respectively. The @Env@ models accessible variables, and @Store@ models
the persistent heap through the program runtime. But the @Addr@ and @Value@ are
just declared as abstract types.

@@

\paragraph{Binding-time Abstraction} As mentioned before, the binding-time is
declared as a higher-kinded type @R[_]@. If we simply instantiate @R@ as an identity
type (i.e., @type R[T] = T@), then the generic interpreter will execute the program.
In Section \ref{stagedinterp}, we will instantiate @R@ using LMS's built-in
next-stage type annotation @Rep@, which makes the interpreter act as a compiler.

\paragraph{Monadic Operations} We define the return type of the interpreter as
@Ans@, which is an arbitrary monad type @AnsM[_]@ wrapping the type @Value@. 
As mentioned in Section ~\ref{monadscala}, to use the @for@-comprehension
syntax, certain constraints have to be posed onto the type @AnsM@. Here, we use a
structural type @MonadOps@ to require @AnsM@ to at least implement @map@ and
@flatMap@. It is worth noting that @MonadOps@ takes another type parameter
@R[_]@ as binding-time; accordingly the outer @R@ defined in the trait is passed
to @MonadOps@. Inside of @MonadOps@, @R[_]@ wraps the data types @A@ and @B@ that are
encapsulated by the monad, but not the monad type @M@ itself. When acting as
compilers, we will also replace the monads to the ones that work on staged
values.

@@

\paragraph{Primitive Operations} Next we define several primitive operations.
Two versions of @alloc@ are declared. The first one takes a store and an
identifier, and produces a fresh address of non-monadic type @R[Addr]@. Since
the freshness of the address may depend on the store, which might be a
next-stage value as indicated by its type, the type of addresses is also wrapped
by @R[_]@. The other one would simply wrap the address with our monadic type
@AnsM[_]@.
\begin{lstlisting}
  def alloc(σ: R[Store], x: Ident): R[Addr];  def alloc(x: Ident): AnsM[Addr]
\end{lstlisting}

Other primitive operations provide basic functionality for the interpreter.
The method @num@ and @close@ deal with primitive values, which lift literal
terms (e.g., lambdas) to our value representation (e.g., closures).
The method @get@ simply retrieves the value mapped from the identifier @x@ in
the environment and store. Conditionals and arithmetics is handled by @br0@
and @arith@, respectively. The methods @ap_clo@ takes a function value and an
argument value and then does the application. Note that the @Env@, @Store@, and
@Value@ are all wrapped by @R[_]@ since they will be known as next-stage value
when acting as compilers.

@@

\paragraph{The Interpreter} Now we can define the semantics-agnostic interpreter
in monadic form. The essential idea is to traverse the abstract syntax tree
while maintaining the effects such as reader and state. 
It is worth noting that the interpreter is written in open-recursive style -- it
can not refer to itself directly, instead, @eval@ takes an additional parameter
@ev@ of type @Expr => Ans@ to refer to itself. Accordingly, the method @close@
that lifts lambda terms to closures and @ap_clo@ that applies functions also
takes an extra @ev@.

@@

\paragraph{Concrete Components}
Now we can instantiate the interpreter as a standard environment-store
interpreter. Such interpreter can be obtained by refunctionalization of CESK
machines \cite{Felleisen:1987:CAH:41625.41654, DBLP:conf/ppdp/AgerBDM03}.
The two types we need to concretize are @Addr@ and @Value@, afterward, the @Env@
and @Store@ is concretized automatically. To assure the freshness, we use @Int@
for the address space @Addr@. A value can be either a tagged number
\texttt{IntV}, or a closure \texttt{CloV} that contains a lambda term and an
environment. The final return value of the interpreter is a pair of values and
stores, where the @Value@ and @Store@ are obviously next-stage objects. We also
provide a standard fixed-point combinator to close the open-recursive function @ev@.

\begin{lstlisting}
  trait ConcreteComponents extends Semantics {
    type Addr = Int
    sealed trait Value
    case class IntV(i: Int) extends Value
    case class CloV(λ: Lam, e: Env) extends Value
    type Result = (R[Value], R[Store])
    def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => ev(fix(ev))(e)
  }
\end{lstlisting}
%\vspace{-1em}

\paragraph{Unstaged Concrete Monads}
% The data manipulated by the monad are staged value. But the monad is known at
% current stage.
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other word, we instantiate the @AnsM@ moand as a stack of @ReaderT@ and @StateT@
transformers, where the @ReaderT@ is parameterized by type @Env@, @StateT@ is
parameterized by type @Store@, and the inner-most monad @IdM@ is just an identity
monad.

\begin{lstlisting}
  trait ConcreteSemantics extends ConcreteComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[IdM, Store, ?], Env, T]
    ...
  }
\end{lstlisting}

Here we sketch the basic idea of unstaged @ReaderT@ and @StateT@. Readers may
refer to \todo{cite, or appendix?} for more detail and implementation of monad
transformers.
A @ReaderT@ monad transformer encapsulates the computation @R => M[A]@, where
@R@ is the reader type, @M[_]@ is a monad type. The @ReaderT@ monad returns the
transformed value of type @M[A]@. Similarly, a @StateT@ monad encapsulates the
computation @S => M[(A, S)]@, where @S@ is the state type, @M[_]@ is a monad
type. The @StateT@ monad returns the transformed value of type @M[(A, S)]@,
where the new state is accompanied.
Note that the binding-time type @R@ is an identity type, thus the monads are
operated on unstaged data. We can also see this from the signature of @flatMap@:
the function @f@ takes an unstaged value of type @A@ and produces a monadic
value.

\begin{lstlisting}
  case class ReaderT[M[_]: Monad, R, A](run: R => M[A]) {
    def flatMap[B](f: A => ReaderT[M, R, B]): ReaderT[M, R, B] = ... }
  case class StateT[M[_]: Monad, S, A](run: S => M[(A, S)]) {
    def flatMap[B](f: A => StateT[M, S, B]): StateT[M, S, B] = ... }
\end{lstlisting}

Then, the monadic operations that manipulate the environment and store are
simply constructing the proper monads and lifting it to the top-level of our
moand stack. To modify the store, for example, we may have a @StateT@ monad that
transforms the current store $\sigma$ to @σ + αv@, which results in
@StateT[IdM, Store, Unit]@, and then lift this @StateT@ value to @ReaderT@,
i.e., @AnsM[Unit]@.
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ + αv))
\end{lstlisting}

\paragraph{Primitive Operations}
Other primitive operations can be implemented straightforwardly, for example,
@alloc@ simply takes the size of current store and produce the successor number.
We elide most of them, but describe a little bit on how do we handle functions
and applications as they will be very different when staging and abstract
interpretation is involved. The method @close@ takes a lambda term and an
environment, then create a closure value @CloV@.

\begin{lstlisting}
  def close(ev: EvalFun)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
\end{lstlisting}

To apply a function, the method @ap_clo@ takes a function value and an arugment
value, extracts the lambda term and environment enclosed in the function value,
and evaluates the body expression of that lambda under the new environment and store.

\begin{lstlisting}
  def ap_clo(ev: EvalFun)(f: Value, arg: Value): Ans = f match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → arg)
      rt <- ext_env(ev(e))(x → α)
    } yield rt
  }
\end{lstlisting}

With other primitive operations implementated, we can implement the top-level
@run@ method, where $\rho_0$ and $\sigma_0$ are initial empty environment and
store.

\begin{lstlisting}
  def run(e: Expr): Result = fix(eval)(e)(ρ$_0$)(σ$_0$)
\end{lstlisting}
